---
title: "MuRD"
output: html_document
vignette: >
  %\VignetteIndexEntry{MuRD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**MuRD** unifies multiple deconvolution schemes to infer cell type proportions from the target bulk RNA-seq data. Three unique features are embraced in this algorithm: first, **MuRD** is able to incorporate extra biological information from external data sources enables (e.g., scRNA_seq and other bulk RNA_seq data from independent studies); second, **MuRD** calibrates the reference-free algorithm by taking into account the proportion estimates from a reference-based approach; third, **MuRD** is robust to incorrect information from any of the provided data sources.

## Installation
```{r}
#package install
library(MuRD)
```

## Data Input
**MuRD** deconvolutes raw read counts data which is in `ExpressionSet` objects based on the package [_Biobase_](https://bioconductor.org/packages/release/bioc/html/Biobase.html). As an illustration of the **MuRD** framework, we conduct both simulation studies and real data analysis based on pancreatic islet data from [Fadista et al. (2014)](https://www.pnas.org/content/111/38/13924) as presented in our paper. Three public scRNA-seq datasets from [Baron et al. (2016)](https://www.sciencedirect.com/science/article/pii/S2405471216302666?via%3Dihub), [Segerstolpe et al. (2016)](https://www.sciencedirect.com/science/article/pii/S1550413116304363?via%3Dihub) and [Xin et al. (2016)](https://www.sciencedirect.com/science/article/pii/S155041311630434X?via%3Dihub) are used. The processed data are available at the ... 

